---
title: "Class Project Machine Learning"
author: "Dr. B"
date: "Friday, December 12, 2014"
output: html_document
---

```{r setup,warning=FALSE,message=FALSE}
##Clear the environment
rm(list=ls())

##Turn off scientific notations for numbers
options(scipen = 999)  

##Set locale
Sys.setlocale("LC_ALL", "English") 

##Load libraries
call <- function(x)
{
        if (!require(x,character.only = TRUE))
        {
                install.packages(x,dep=TRUE)
        }
}

call("caret")
call("randomForest")

##Set seed for reproducibility
set.seed(2345)

##Set location of Data DOIrectory
datadir=('C:/Users/bryan_000/Documents/GitHub/Data/')
```

##Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

(More information is available from the website [here](http://groupware.les.inf.puc-rio.br/har). (see the section on the Weight Lifting Exercise Dataset)). 

###Load the Data
The data for this project come from this [source](http://groupware.les.inf.puc-rio.br/har). The training data for this project were downloaded from [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv).  The test data were downloaded from [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv). 


```{r load}
#Set name of training and testing datafiles
train.data=paste(datadir,"pml-training.csv",sep = "")
test.data=paste(datadir,"pml-testing.csv",sep = "")
train.url="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

##Check for training file, download if not there
if (!file.exists(train.data)) {
  download.file(train.url, destfile = train.data)
}

##Check for testing file, download if not there
if (!file.exists(test.data)) {
  download.file(test.url, destfile = test.data)
}

##Read the data
test <- read.csv(test.data, sep = ",", na.strings = c("", "NA"))
train <- read.csv(train.data, sep = ",", na.strings = c("", "NA"))

##Clean out variables not used
rm(test.data)
rm(train.data)
rm(test.url)
rm(train.url)
rm(datadir)
```

###Clean the Data
Once the data were loaded, the next step was to clean it by removing columns full of NAs and removing features that were not in the test set. Because the test dataset had no time-dependence, these values were disregarded from the training dataset. The first 7 features were removed because they were related to the time-series or were not numeric.

```{r clean}
##Clean the data
names(test) <- tolower(names(test))
names(test) <- gsub("_","",names(test))
names(train) <- tolower(names(train))
names(train) <- gsub("_","",names(train))

##Remove columns full of NAs. Only use features used in test dataset.
features <- names(test[,colSums(is.na(test)) == 0])[8:59]
train.df <- train[,c(features,"classe")]
test.df <- test[,c(features,"problemid")]

##clean out unused datasets
rm(test)
rm(train)
```
